On the Limits of Making Programming Easy

### 2024-12-01 paper.tex
Easiest and other changes

\title{On the limits of making programming easy}
\titlerunning{On the limits of making programming easy}

@@

\begin{abstract}
A lot of programming research shares the same basic motivation: how can we make programming easier? Alas, this problem is difficult to tackle directly. Programming is a tangle of conceptual models, programming languages, user interfaces and more and we cannot advance all of these at the same time. Moreover, we have no good metric for measuring whether programming is easy. As a result, we usualy give up on the original motivation and pursue narrow tracktable research for which there is a rigorous methodology.

@@

\section{Introduction}
It is a cold winter morning on 1st December 2023 outside the Cambridge Computer Laboratory. In the area known as The Street in front of the lecture theaters, on the cold winter day of 1st December 2023, we encounter two computer scientists who arrived for the MycroftFest symposium. They are Tomas Petricek, a former PhD student of Alan, and Joel Jakubovic, who is Tomas' PhD student at the time. Joel recalls how he enjoyed Alan's \emph{Concepts in Programming Languages} module and the two start discussing their research.

@@

\says[\T]{My PhD was about making programming easier too. I worked on client-server web programming before~\cite{petricek-2007-fswebtools} and found it tricky to keep track of the right context on the two sites. Some code requires resources that are only available on the server (e.g. a database), while some requires resources only available in the web browser (e.g., the user's location).}

@@

\vspace{0.5em}
(b) A screenshot of Hypercard, an early hyper-text system that allowed \\
users to easily create their own interactive ``decks''.
\vspace{0.5em}

@@

\says[\J]{First, the different view lets us consider ways to make programming easier which are hard to think about as languages. For example, visual programming, block-based editing \cite{resnick-2009-scratch} or systems like Flash \cite{ankerson-2018-dot} and Hypercard. These are programming systems we should learn from. Second, the view lets us talk about interactivity and self-sustainable systems that can be modified from within themselves \cite{hirschfeld-2008-s3,hirschfeld-2010-s3}. It would be useful to understand image-based programming systems like Lisp and Smalltalk, because they can in principle be used to make their \emph{own} programming easier.}

@@

\says[\T]{Brooks \cite{brooks-1987-silver} analysed the different kinds of complexity involved in building software. The \emph{essential complexity} is inherent in the real-world. It comes from the fact that our software has to encode the many (perhaps ad-hoc) rules of the human world. The \emph{accidental compexity} comes from imperfection in our tools and languages.}

@@

\begin{figure}[t]
\centering
\includegraphics[scale=0.33]{figures/mit20.jpg}
\caption{MIT Building 20 was a temporary timber building, erected for radar research during the World War II. It continued to exist for over 50 years and was highly popular for its infinite
adaptability \cite{brand-1994-buildings} (foto: MIT Museum).}
% https://www.ll.mit.edu/about/history/mit-radiation-laboratory
\label{fig:mit20}
\end{figure}

@@

\says[\T]{You are probably right. Buildings seem like the most static things, yet they evolve. Software is more dynamic, so it must surely evolve too. But how can this solve the problem of software complexity?}

@@

\says[\J]{Well, a programming system should model the chain of cause to effect in programs that you create. Then, when you want to make a change to something that you can see, or to something that produces visible outputs or effects on the screen, it can point you back to the code that you need to change.}

\says[\T]{Well, I'm not sure what the programmer interaction for this could look like, but I think I can imagine the technical characteristics of a system necessary to make it work. The system would need to record information about its execution and make it acessible. Basman \cite{basman-2022-materialized} refers to this idea as ``materialized execution'' and traces it to the 1980s Boxer project \cite{disessa-1986-boxer}. A similar idea was also used recently to automatically link data visualizations by analysing code used to produce them \cite{perera-2022-galois}.}

\says[\J]{This is useful technical background, but we also need to look to interesting programming systems for inspirations about the interface through which we can make such program editing accessible.}

@@

\says[\J]{Yes, those tools give you rapid feedback. Perhaps a surprisingly good example here is Excel, which also, to some extent, keeps track of the chain of cause to effect in that you can view the source of the formula and see what other cells were used to compute the result. You do have to backtrace the causal chain manually by looking up the cells one by one. But it is the best example I can think of.}

\says[\T]{Ah, but this only works with the relatively simple formula language. If you want to achieve something more complicated, you have to do that with VBA macros, which is as complicated as regular programming. }

\begin{figure}[t]
\centering
\begin{subfigure}{0.48\textwidth}
   \noindent
   \includegraphics[scale=0.65]{figures/sub1.pdf}
   \caption{Excel offers multiple but disconnected substrates (data entry, formulas, macros) than a reguler end-user application.}
   \label{fig:sub1}
\end{subfigure}
\hfill
\begin{subfigure}{0.48\textwidth}
   \noindent
   \includegraphics[scale=0.65]{figures/sub2.pdf}
   \caption{An ideal system would have one adaptable substrate that makes minimal changes simple, but allows harder larger chages.}
   \label{fig:sub2}
\end{subfigure}
\caption{Scope of change vs. difficulty of use of multiple programming substrates.}
\label{fig:substrates}
\end{figure}

@@

\says[\T]{There is perhaps a nice characterization of the ideal notational system that we are looking for in the work of Pierre Depaz \cite[p.123]{depaz-2023-aesthetics} on aesthetics of code. He says that \emph{``simplicity is found in source code when the syntax and the ontologies used are an exact fit to the problem: simple code is code that is neither too precise, nor too generic, displaying an understanding of and a focus on the problem domain, rather than the applied tools.''}}

\says[\J]{I think this really boils down to two things. First, you need notational pluralism to be able to use the right tools for different jobs, and second, non-textual notations should be on the menu, because those are sometimes the most appropriate fit for a problem.}

\says[\T]{Hmm, but isn't notational pluralism impractical? In Excel, you have multiple disconnected notations such as tables, formulas and macros. But this also means that there is no gradual progression between them. If you learn how to use formulas, it does not help you at all with learning how to write macros (or even become a developer working on Excel at Microsoft...).}

@@

\says[\J]{I imagine it could work in spreadsheets. You could have more things in Excel modifiable in a table. The key/value pairs in the preferences menu could, in addition to their normal UI, be brought into the table substrate. More of the interface could also be defined in a table. And then you could also have your preferences computed with formulas!}

@@

\says[\T]{Well, to take this idea to its limits, are you saying that MS Paint should be written in the Piet esolang that we can see in Figure~\ref{fig:piet}?!}

\says[\J]{Fair enough. The principle breaks down with absurd examples if you take it literally, but that doesn't mean the principle will never work. It is still useful, but you need to think about what really is the essence of the substrate through which most people interact with computers today. It is not pixels...}

@@

\says[\J]{I do not have a perfect answer to this question, but you can think of the basic substrate as a sort of ``user interface physics'' within which specific visual notations for particular problems could be defined.\footnote{Constraint-based systems for specifying user interfaces, such as Cassowary \cite{borning-1997-constraints}, can be seen as related to this idea, although they focus more on rendering the layout for the end-user than on developing a foundational underlying substrate. Additionally, individual constraints are too low-level for a human-friendly UI substrate, although they could well be what such a substrate ``compiles'' to in order to run.} If we had that, then I think we would make a step towards the ideal substrate for programming. And the substrate needs to live within an actual running materialized system, rather than a static pre-scription of it like a program in source code today.}

@@

\says[\T]{Well, you remember correctly. There was a point in my PhD when I realized that all the things I could prove about my programming langauge were not that interesting and all the things that I found interesting about programming were not particularly formalizable or provable. I guess I was getting increasingly interested in what people sometimes call ``programming experience'' research.\footnote{The topic is the subject of the Programming Experience workshop\\ (https://programming-experience.org) that has been running since 2016.} The experience also made me think much more about research methodology and led me to interest in history and philosophy of science.}

@@

\says[\T]{Well, for one thing, there is a bias for theory over experiment in all branches of science. Ian Hacking \cite{hacking-1983-representing} pointed this out in the case of physics and the same applies to computer science. However, finding a suitable methodology for programming experience research is an open problem that does not have a definite answer. It may need a range of qualitative and quantitative methods, including formalisms and empirical evaluation, but also system descriptions, user studies and perhaps even interactive artifacts \cite{edwards-2019-eval}.}

@@

\says[\J]{Yes, it would be a corrective to the shift that Gabriel observed in the opposite direction\ \cite{rpg-2012-revolution}; not by \emph{fighting} PL research, but by simply making space to do the more general thing. To truly establish a new paradigm, then the shift to thinking about systems would put more emphasis on research methods that can talk about them in their full complexity. It would not \emph{invalidate} programming language theory, but it could make it more of a speciality instead of the default. However, we are quite far from what Kuhn called ``normal science''. It is still early days for programming systems research!}

@@

\says[\J]{Right. The full realisation of the aims of the framework is a long-term research program. What we have already is the seed to start it off. We have a set of dimensions that come with qualitative descriptions. Future work, as set out towards the end of my dissertation, involves making these descriptions more precise and possibly even quantifiable.}

@@

\says[\T]{My tongue-in-cheek suggestion is that your
\emph{Concepts in Programming Languages} module should become
\emph{Concepts in Programming Systems}! We need a new generation of programming systems researchers who will adopt this new way of thinking about programming and come up with great new theories and experimental systems!}

>>>>

\title{On the Limits of Making Programming Easy}
\titlerunning{On the Limits of Making Programming Easy}

@@

\begin{abstract}
A lot of programming research shares the same basic motivation: how can we make programming easier? Alas, this problem is difficult to tackle directly. Programming is a tangle of conceptual models, programming languages, user interfaces and more and we cannot advance all of these at the same time. Moreover, we have no good metric for measuring whether programming is easy. As a result, we usualy give up on the original motivation and pursue narrow tractable research for which there is a rigorous methodology.

@@

\section{Introduction}
A cold winter morning, 1st December 2023, outside the Cambridge Computer Laboratory; two computer scientists arrive in the area known as The Street in front of the lecture theaters; they aim to attend the MycroftFest symposium: Tomas Petricek---a former PhD student of Alan---and Joel Jakubovic, Tomas’ PhD student. The elder and younger researchers discuss threads of programming study as different from each other as The Street and the Cambridge Computer Laboratory are from Trinity Street and Peterhouse.

Let's eavesdrop.

@@

\says[\T]{My PhD was about making programming easier too. I worked on client-server web programming before~\cite{petricek-2007-fswebtools} and found it tricky to keep track of the right context on the two sites. Some code requires resources that are only available on the server (e.g. a database), while some requires resources only available in the web browser (e.g. the user's location).}

@@

\vspace{0.5em}
(b) A screenshot of Hypercard, an early hypertext system that allowed \\
users to easily create their own interactive ``decks''.
\vspace{0.5em}

@@

\says[\J]{First, the different view lets us consider ways to make programming easier which are hard to think about as languages. For example, visual programming, block-based editing \cite{resnick-2009-scratch} or systems like Flash \cite{ankerson-2018-dot} and Hypercard. These are programming systems we should learn from. Second, the view lets us talk about interactivity and self-sustainable systems that can be modified from within themselves \cite{hirschfeld-2008-s3,hirschfeld-2010-s3}. It would be useful to understand highly malleable programming systems like Lisp and Smalltalk, because they can in principle be used to make their \emph{own} programming easier.}

@@

\says[\T]{Brooks \cite{brooks-1987-silver} analysed the different kinds of complexity involved in building software. The \emph{essential complexity} is inherent in the real-world. It comes from the fact that our software has to encode the many (perhaps ad-hoc) rules of the human world. The \emph{accidental complexity} comes from imperfection in our tools and languages.}

@@

\begin{figure}[t]
\centering
\includegraphics[scale=0.33]{figures/mit20.jpg}
\caption{MIT Building 20 was a temporary timber building, erected for radar research during the World War II. It continued to exist for over 50 years and was highly popular for its infinite
adaptability \cite{brand-1994-buildings} (photo: MIT Museum).}
% https://www.ll.mit.edu/about/history/mit-radiation-laboratory
\label{fig:mit20}
\end{figure}

@@

\says[\T]{You are probably right. Buildings seem like the most static things, yet they evolve. Software, being structured information, ought to be much more malleable than the rigid materials of buildings; accordingly, it evolves much faster. But how can this solve the problem of software complexity?}

@@

\says[\J]{Well, a programming system should model the chain of cause to effect in programs that you create. Then, when you want to make a change to something that you can see, or to something that produces visible outputs or effects on the screen, it can point you back to the code that you need to change. Bret Victor demonstrated what this could look like \cite{bret-victor}, but I'm unaware of any follow-up work to realise his vision---which raises the question of how he himself got his demo to work, something that does occasionally keep me up at night.}

\says[\T]{I think I can imagine the technical characteristics such a system would need to work: it would need to record information about its execution and make it acessible. Basman \cite{basman-2022-materialized} refers to this idea as ``materialized execution'' and traces it to the 1980s Boxer project \cite{disessa-1986-boxer}. A similar idea was also used recently to automatically link data visualizations by analysing code used to produce them \cite{perera-2022-galois}.}

\says[\J]{This is useful technical background, but we should also look to interesting \emph{real} programming systems---not just demos---for inspiration about interfaces that would make such program editing accessible.}

@@

\says[\J]{Yes, those tools give you rapid feedback. Perhaps a surprisingly good example here is Excel, which also, to some extent, keeps track of the chain of cause to effect in that you can view the source of the formula and see which other cells were used to compute the result. You do have to backtrace the causal chain manually by looking up the cells one by one. But it is the best example I can think of.}

\says[\T]{In fact, Excel has pretty good backtracing support now! But regardless of whether you trace manually or with the system's help, this only works with the relatively simple formula language. If you want to achieve some more complicated behaviour, you have to do that with VBA macros and give up this traceability. }

\begin{figure}[t]
\centering
\begin{subfigure}{0.48\textwidth}
   \noindent
   \includegraphics[scale=0.65]{figures/sub1.pdf}
   \caption{Excel offers multiple but disconnected substrates (data entry, formulas, macros).}
   \label{fig:sub1}
\end{subfigure}
\hfill
\begin{subfigure}{0.48\textwidth}
   \noindent
   \includegraphics[scale=0.65]{figures/sub2.pdf}
   \caption{An ideal system would have one adaptable substrate that makes minimal changes simple, but allows harder larger chages.}
   \label{fig:sub2}
\end{subfigure}
\caption{Scope of change vs. difficulty of use of multiple programming substrates.}
\label{fig:substrates}
\end{figure}

@@

\says[\T]{There is perhaps a nice characterization of the ideal notational continuum that we are looking for in the work of Pierre Depaz \cite[p.123]{depaz-2023-aesthetics} on aesthetics of code. He says that \emph{``simplicity is found in source code when the syntax and the ontologies used are an exact fit to the problem: simple code is code that is neither too precise, nor too generic, displaying an understanding of and a focus on the problem domain, rather than the applied tools.''}}

\says[\J]{I think this really boils down to two things. First, to be able to use the right tools for different jobs, you need to be free from a restriction to just one notation. Second, you should also be free from a restriction to plain text; non-textual notations should be on the menu, because those are sometimes the most appropriate fit for a problem.}

\says[\T]{Hmm, but wouldn't such a ``notational pluralism'' be impractical? In Excel, you may use multiple disjoint notations such as tables, formulas and macros---yet this also means that there is no gradual progression between them. If you learn how to use formulas, it does not help you at all with learning how to write macros (or even become a developer working on Excel at Microsoft...).}

@@

\says[\J]{Something like this could work in spreadsheets. I can think of some parts of Excel-the-application that transfer well to a portion of a spreadsheet. One could press a button to ``reflect'' the key/value pairs in the preferences menu into two columns of cells. When edited, these cells would update Excel's internal preferences that you normally edit in a dialog box. At a stretch, perhaps other parts of such a dialog could also be reflected---say, the list of options in a drop-down box. And then you could, for example, have your preferences computed with formulas!}

@@

\says[\T]{Well, this is a heavy constraint on the notations we can use for development! What would justify such an idea?}

\says[\J]{I admit, it's about as restrictive as one could get! The motivation is that the user is already familiar with the ``notation for use'' in the software---so to the extent that ``development'' can be massaged to use the same notation, they are able to adapt the software by re-using the same skills. They no longer have to invest in learning a completely different notation (like a programming language) before they can do that.}

\says[\T]{So, to take this idea to its limits, you're saying that MS Paint should be written in the Piet esolang that we can see in Figure~\ref{fig:piet}?!}

\says[\J]{Fair enough. I wouldn't go so far; *given that* the use-notation in Paint is what it is, there aren't many changes to Paint that fit sensibly within it. That being said---the program's tool icons could be ``reflected'' in its canvas and re-drawn that way, but that's the only straw I can think to grasp! The principle will break down with absurd examples if you commit to it fully, but that doesn't mean we shouldn't explore its domain of validity. It is still promising, but you need to think about what really is the essence of the substrate through which most people interact with computers today. It is not pixels...}

@@

\says[\J]{I do not have a perfect answer to this question, but you can think of the basic substrate as a sort of ``user interface physics'' within which specific visual notations for particular problems could be defined.\footnote{Constraint-based systems for specifying user interfaces, such as Cassowary \cite{borning-1997-constraints}, can be seen as related to this idea, although they focus more on rendering the layout for the end-user than on developing a foundational underlying substrate. Additionally, individual constraints are too low-level for a human-friendly UI substrate, although they could well be what such a substrate ``compiles'' to in order to run. This was the subject of much research in the 1990s \cite{myers-1992-uilangs}.} Such a substrate could even be the result of taking a 2D physics engine, stripping it of concepts like mass and rotations, and optimising it for this simpler domain. If we had that, then I think we would make a step towards the ideal substrate for programming. And the substrate would need to live within an actual running materialized system, rather than a static blueprint like a program in source code today.}

@@

\says[\T]{Well, you remember correctly. There was a point in my PhD when I realized that all the things I could prove about my programming language were not that interesting, and all the things that I found interesting about programming were not particularly formalizable or provable. I guess I was getting increasingly interested in what people sometimes call ``programming experience'' research.\footnote{The topic is the subject of the Programming Experience workshop\\ (https://programming-experience.org) that has been running since 2016.} The experience also made me think much more about research methodology and led me to interest in history and philosophy of science.}

@@

\says[\T]{Well, for one thing, there is a bias for theory over experiment in all branches of science. Ian Hacking \cite{hacking-1983-representing} pointed this out in the case of physics, as did Hossenfelder \cite{2018-hossenfelder-lost}, and the same applies to computer science. However, finding a suitable methodology for programming experience research is an open problem that does not have a definite answer. It may need a range of qualitative and quantitative methods, including formalisms and empirical evaluation, but also system descriptions, user studies and perhaps even interactive artifacts \cite{edwards-2019-eval}.}

@@

\says[\J]{Yes, it would be a corrective to the shift that Gabriel observed in the opposite direction\ \cite{rpg-2012-revolution}; not by \emph{fighting} PL research, but by simply making space to do the more general thing. To truly establish a new paradigm, the return to thinking about systems would put more emphasis on research methods that can talk about them in their full complexity. It would not \emph{invalidate} programming language theory, but it could make it more of a speciality instead of the default. However, we are quite far from what Kuhn called ``normal science''. It is still early days for programming systems research!}

@@

\says[\J]{Right. The full realisation of the aims of the framework is a long-term research program. What we have already is the seed to start it off. We have a set of dimensions that come with qualitative descriptions. Future work, as set out towards the end of my dissertation \cite{jakubovic-2024-phd}, involves making these descriptions more precise and possibly even quantifiable.}

@@

\says[\T]{My tongue-in-cheek suggestion is that your
\emph{Concepts in Programming Languages} module, which I very much enjoyed back in the day, should become
\emph{Concepts in Programming Systems}! We need a new generation of programming systems researchers who will adopt this new way of thinking about programming and come up with great new theories and experimental systems!}
